### 2020-01-30 S2_appendix.tex
Interval operator

\begin{figure}[!htb]
\includegraphics[width=0.85\textwidth]{Figures/CalSim_real.pdf}
\caption{\textbf{The \textit{real} parameterisation,} with the operators decribed by Zhang and Drummond 2020.}
\label{fig:rateparams}
\end{figure}

\begin{figure}[!htb]
\includegraphics[width=0.85\textwidth]{Figures/CalSim_quant.pdf}
\caption{\textbf{The \textit{quant} parameterisation,} with the operators decribed by Zhang and Drummond 2020, and adapted for \textit{quant} as described in \textbf{S1 Appendix}.}
\label{fig:rateparams}
\end{figure}

>>>>

\begin{figure}[!htb]
\includegraphics[width=0.85\textwidth]{Figures/CalSim_real.pdf}
\caption{\textbf{The \textit{real} parameterisation,} with the operators decribed by Zhang and Drummond 2020. The \textit{real 2006} setting uses a subset of the operators tested in this validation. }
\label{fig:rateparams}
\end{figure}

\begin{figure}[!htb]
\includegraphics[width=0.85\textwidth]{Figures/CalSim_quant.pdf}
\caption{\textbf{The \textit{quant} parameterisation,} with the operators decribed by Zhang and Drummond 2020, and adapted for \textit{quant} as described in \textbf{S1 Appendix}. The \textit{quant 2006} setting uses a subset of the operators tested in this validation.}
\label{fig:rateparams}
\end{figure}

\begin{figure}[!htb]
\includegraphics[width=0.85\textwidth]{Figures/CalSim_NER_real.pdf}
\caption{\textbf{The Narrow Exchange Rate operators under the \textit{real} parameterisation.} The 48 NER and 48 NERw operators, including the null Narrow Exchange operator, are tested in the same MCMC session.}
\label{fig:rateparams}
\end{figure}

\begin{figure}[!htb]
\includegraphics[width=0.85\textwidth]{Figures/CalSim_NER_quant.pdf}
\caption{\textbf{The Narrow Exchange Rate operators under the \textit{quant} parameterisation.} The 48 NER and 48 NERw operators are tested simultaneously.}
\label{fig:rateparams}
\end{figure}

\begin{figure}[!htb]
%\includegraphics[width=0.85\textwidth]{Figures/CalSim_quant.pdf}
\caption{\textbf{Bactrian proposal kernel.} A $\text{Bactrian}(0.95)$ proposal kernel is used for all clock model operators, as described in main article.}
\end{figure}

### 2020-01-30 manuscript.tex
Interval operator

The step size of a proposal kernal $q(x^\prime|x)$ should be such that the proposed state $x^\prime$ is sufficiently far from the current state $x$ to explore vast areas of parameter space, but not so large that the proposal is rejected too often. Yang et al. have challenged the widely used uniform proposal kernel in place of the Bactrian kernel \cite{yang2013searching, thawornwattana2018designing}.
The $\text{Bactrian}(m)$ distribution is defined as the sum of two Normal distributions:

@@

In this article we compare the performance the uniform and Bactrian proposal kernels in the clock model. Three Bactrian distributions are compared($m=0.92$, $m=0.95$, and $m=0.98$). The clock model operators which are augmented by the Bactrian distribution are described in \textbf{Table \ref{table:kernels}}.

@@

\begin{table}[h!]
\centering
\begin{tabular}{|l| l l l|} 
 \hline
 & Operator & Proposal & Parameters  \\
 1 & \textbf{Random walk operator} & $x^\prime \leftarrow x + s\Sigma$ & $t, \sigma, r, q$ \\
 2 & \textbf{Scale operator} & $x^\prime \leftarrow x \times e^{s\Sigma}$ & $\sigma, r$  \\
 %3 & \textbf{Uniform operator} & $\begin{array} {rl} y &\leftarrow s\Sigma \times \frac{u - x}{x - l} \\ x^\prime &\leftarrow \frac{u + l * y}{y + 1}  \end{array}$ & $t, q$  \\
 3 & \textbf{Constant distance operators} & $x^\prime \leftarrow x + s\Sigma$ & $t$ \\
 \hline
\end{tabular}
\caption{Summary of proposal kernels $q(x^\prime|x)$ of clock model operators. In each operator, $\Sigma$ is drawn from either a $\text{Noramlised-Bactrian}(m)$ or $\text{Uniform}(-\sqrt{3},\sqrt{3})$ distribution. The scale size $s$ is tunable. The proposal kernel may apply to node heights $t$, clock standard deviation $\sigma$, clock rates $r$ (\textit{real} only), and clock rate quantiles $q$ (\textit{quant} only). The Scale operator acts on parameters with non-negative domains. }
\label{table:kernels}
\end{table}

@@

Lakner et al. 2008 \cite{lakner2008efficiency} found that tree operators which perterb topology (such as NNI and SPR) consistently perform better than those which also change branch lengths (such as LOCAL \cite{simon1998local} and Continuous Change \cite{jow2002bayesian}). If Narrow Exchange was adapted to the relaxed clock model by ensuring that genetic distances remain constant after the proposal, its performance may be improved even further, which may in turn permit proposing a new node height $t_D$ and therefore changing branch time lengths (without affecting branch distance lengths).

@@

Here, we present the \textbf{Narrow Exchange Rate} (NER) operator. Let $r_A$, $r_B$, $r_C$, and $r_D$ be the clock rates of nodes $A$, $B$, $C$, and $D$, respectively. In addition to the modest topological change applied by Narrow Exchange, NER also proposes new clock rates ${r_A}^\prime$, ${r_B}^\prime$, ${r_C}^\prime$, and ${r_D}^\prime$. We also consider a case of the NER operator which proposes a new node height:

\begin{align}
	{t_D}^\prime \leftarrow t_D + \Sigma
\end{align}

for random walk step size $\Sigma$ drawn from a uniform or \textbf{\nameref{sect:randomwalks}}. NER is compatible with both the \textit{real} and \textit{quant} parameterisations. Analagous to the Constant Distance operator, these rates are proposed such that genetic distances between nodes $A$, $B$, $C$, and $E$ are maintained. Thus, there are $\binom{4}{2} = 6$ pairwise distance constraints.

\begin{align}
	\mathcal{D}_{AB}: \quad & r_A  (t_D - t_A) + r_B  (t_D - t_B) = \nonumber \\
					 & {r_A}^\prime  (t_E - t_A) + {r_D}^\prime  (t_E - t_D) + {r_B}^\prime (t_D - t_B) \\
	\mathcal{D}_{AC}: \quad & r_A  (t_D - t_A) + r_D  (t_E - t_D) + r_C  (t_E - t_C) = \nonumber \\
				 	  & {r_A}^\prime  (t_E - t_A) + {r_D}^\prime  (t_E - t_D) + {r_C}^\prime (t_D - t_C) \\
 	\mathcal{D}_{AE}: \quad & r_A  (t_D - t_A) + r_D  (t_E - t_D)= \nonumber \\
					  & {r_A}^\prime  (t_E - t_A) \\
  	\mathcal{D}_{BC}: \quad & r_B  (t_D - t_B) + r_D  (t_E - t_D) + r_C  (t_E - t_D)= \nonumber \\
					  & {r_B}^\prime (t_D - t_B) + {r_C}^\prime (t_D - t_C) \\
   	\mathcal{D}_{BE}: \quad & r_B  (t_D - t_B) + r_D  (t_E - t_D)= \nonumber \\
					  & {r_B}^\prime (t_D - t_B) + {r_D}^\prime (t_E - t_D) \\
	\mathcal{D}_{CE}: \quad & r_C  (t_E - t_C)= \nonumber \\
					  & {r_C}^\prime (t_D - t_C) + {r_D}^\prime (t_E - t_D) 
\end{align}

@@

\begin{align}
	r_i > 0 \text{ and } {r_i}^\prime > 0 \text { for } i \in \{A,B,C,D\}.
\end{align}

Unfortunately, it is not possible to solve all six $\mathcal{D}_{ij}$ constraints without permitting non-positive rates. Therefore rather than conserving all six pairwise distances, NER conserves a \textit{subset} of distances. It is not immediately clear which distances should be conserved.

@@

@@

For the latter two criteria, methodologies are benchmarked using 30 empirical datasets compiled and partitioned \cite{lanfear2016partitionfinder} by Lanfear \cite{lanfear2019Github} In order to explore the effects of sequence length, the partitions of each empirical dataset are randomly subsampled at varying sample sizes (\textbf{Table \ref{table:datasets}}). Datasets and BEAST2 .xml templates are provided in the GitHub repository accompanying this article.

@@

% https://github.com/roblanf/BenchmarkAlignments
\begin{table}[h!]
\centering
\begin{tabular}{|l| l l l l l|} 
 \hline
  &&&&& \\
 ID & $N$ & $P$ & $L$ (kb) & $L_\text{eff}$ (kb) & Reference \\
 &&&&& \\

@@

26  &  608  &  1/4/8/10  &  0.9/3.7/7.1/8.6  &  0.7/2.8/5.4/6.4  &  Near 2013 \cite{Near_2013} \\

 &&&&& \\
 \hline
\end{tabular}
\caption{Empirical datasets used during benchmarking. Number of taxa $N$, number of partitions $P$, total alignment length $L$, and number of patterns $L_\text{eff}$ are specified. Where a dataset is sampled more than once, the dimensions of its multiple samples are separated by `/'.  }
\label{table:datasets}
\end{table}

>>>>

The step size of a proposal kernal $q(x^\prime|x)$ should be such that the proposed state $x^\prime$ is sufficiently far from the current state $x$ to explore vast areas of parameter space, but not so large that the proposal is rejected too often \cite{roberts1997weak}. Yang et al. have challenged the widely used uniform proposal kernel in place of the Bactrian kernel \cite{yang2013searching, thawornwattana2018designing}.
The $\text{Bactrian}(m)$ distribution is defined as the sum of two Normal distributions:

@@

In this article we compare the performance of uniform and Bactrian proposal kernels in the clock model. Two Bactrian distributions are compared($m=0.95$ and $m=0.98$). The clock model operators which these proposal kernels apply to are described in \textbf{Table \ref{table:kernels}}.

@@

\begin{table}[h!]
\centering
\begin{tabular}{|l| l l l|} 
 \hline
 & Operator & Proposal & Parameters  \\
 1 & \textbf{Random walk operator} & $x^\prime \leftarrow x + s\Sigma$ & $\sigma, r, q$ \\
 2 & \textbf{Scale operator} & $x^\prime \leftarrow x \times e^{s\Sigma}$ & $\sigma, r$  \\
 3 & \textbf{Interval operator} & $\begin{array} {rl} y &\leftarrow \frac{u - x}{x - l} \times e^{s\Sigma} \\ x^\prime &\leftarrow \frac{u + l * y}{y + 1}  \end{array}$ & $q \;\; (l=0,u=1)$  \\
 4 & \textbf{Constant distance operators} & $x^\prime \leftarrow x + s\Sigma$ & $t$ \\

 \hline
\end{tabular}
\caption{Summary of proposal kernels $q(x^\prime|x)$ of clock model operators. In each operator, $\Sigma$ is drawn from either a $\text{Bactrian}(m)$ or $\text{Uniform}$ distribution (distributions are normalised so that they have a mean of 0 and a variance of 1). The scale size $s$ is tunable. The proposal kernel may apply to node heights $t$, clock standard deviation $\sigma$, clock rates $r$ (\textit{real} only), and clock rate quantiles $q$ (\textit{quant} only). The Scale operator acts on parameters with non-negative domains. The Interval operator proposes values which respect its domain ie. $l < x^\prime < u$. }
\label{table:kernels}
\end{table}

@@

Lakner et al. 2008 \cite{lakner2008efficiency} found that tree operators which perterb topology (such as NNI and SPR) consistently perform better than those which also change branch lengths (such as LOCAL \cite{simon1998local} and Continuous Change \cite{jow2002bayesian}). If Narrow Exchange was adapted to the relaxed clock model by ensuring that genetic distances remain constant after the proposal, its performance may be improved even further. This may in turn permit proposing a new node height $t_D$ and therefore changing branch (time) lengths.

@@

Here, we present the \textbf{Narrow Exchange Rate} (NER) operator. Let $r_A$, $r_B$, $r_C$, and $r_D$ be the clock rates of nodes $A$, $B$, $C$, and $D$, respectively. In addition to the modest topological change applied by Narrow Exchange, NER also proposes new clock rates ${r_A}^\prime$, ${r_B}^\prime$, ${r_C}^\prime$, and ${r_D}^\prime$. While NER does not alter $t_D$ (ie. ${t_D}^\prime \leftarrow t_D$), we also consider NERw - a special case of the NER operator which embarks $t_D$ on a random walk:

\begin{align}
	{t_D}^\prime \leftarrow t_D + s\Sigma
\end{align}

for random walk step size $s\Sigma$ where $s$ is a tunable scaler parameter and $\Sigma$ is drawn from a uniform or \textbf{\nameref{sect:randomwalks}}. NER (and NERw) are compatible with both the \textit{real} and \textit{quant} parameterisations. Analagous to the Constant Distance operator, new rates are proposed such that genetic distances between nodes $A$, $B$, $C$, and $E$ are maintained. Thus, there are $\binom{4}{2} = 6$ pairwise distance constraints.

\begin{align}
	\mathcal{D}_{AB}: \quad & r_A  (t_D - t_A) + r_B  (t_D - t_B) = \nonumber \\
					 & {r_A}^\prime  (t_E - t_A) + {r_D}^\prime  (t_E - {t_D}^\prime) + {r_B}^\prime ({t_D}^\prime - t_B) \\
	\mathcal{D}_{AC}: \quad & r_A  (t_D - t_A) + r_D  (t_E - t_D) + r_C  (t_E - t_C) = \nonumber \\
				 	  & {r_A}^\prime  (t_E - t_A) + {r_D}^\prime  (t_E - {t_D}^\prime) + {r_C}^\prime ({t_D}^\prime - t_C) \\
 	\mathcal{D}_{AE}: \quad & r_A  (t_D - t_A) + r_D  (t_E - t_D)= \nonumber \\
					  & {r_A}^\prime  (t_E - t_A) \\
  	\mathcal{D}_{BC}: \quad & r_B  (t_D - t_B) + r_D  (t_E - t_D) + r_C  (t_E - t_D)= \nonumber \\
					  & {r_B}^\prime ({t_D}^\prime - t_B) + {r_C}^\prime ({t_D}^\prime - t_C) \\
   	\mathcal{D}_{BE}: \quad & r_B  (t_D - t_B) + r_D  (t_E - t_D)= \nonumber \\
					  & {r_B}^\prime ({t_D}^\prime - t_B) + {r_D}^\prime (t_E - {t_D}^\prime) \\
	\mathcal{D}_{CE}: \quad & r_C  (t_E - t_C)= \nonumber \\
					  & {r_C}^\prime ({t_D}^\prime - t_C) + {r_D}^\prime (t_E - {t_D}^\prime) 
\end{align}

@@

\begin{align}
	r_i > 0 \text{ and } {r_i}^\prime > 0 \text { for } i \in \{A,B,C,D\} \\
	\max\{t_B, t_C \} < {t_D}^\prime < t_E.
\end{align}

Unfortunately, it is not possible to solve all six $\mathcal{D}_{ij}$ constraints without permitting non-positive rates or illegal trees. Therefore rather than conserving all six pairwise distances, NER conserves a \textit{subset} of distances. It is not immediately clear which distances should be conserved.

@@

% Cite HKY somewhere hasegawa1985dating

@@

For the latter two criteria, methodologies are benchmarked using 25 empirical datasets compiled and partitioned \cite{lanfear2016partitionfinder} by Lanfear \cite{lanfear2019Github} In order to explore the effects of sequence length, the partitions of each empirical dataset are randomly subsampled at varying sample sizes (\textbf{Table \ref{table:datasets}}). Datasets and BEAST2 .xml templates are provided in the GitHub repository accompanying this article.

@@

% https://github.com/roblanf/BenchmarkAlignments
\begin{table}[h!]
\centering
\begin{tabular}{|l| l l l l l|} 
 \hline
  & $N$ & $P$ & $L$ (kb) & $L_\text{eff}$ (kb) & Ref. \\

@@

 \hline
\end{tabular}
\caption{Empirical datasets used during benchmarking. Number of taxa $N$, number of partitions $P$, total alignment length $L$, and number of patterns $L_\text{eff}$ are specified. Where a dataset is sampled more than once, the dimensions of its multiple samples are separated by `/'.  }
\label{table:datasets}
\end{table}
